---
title: "506CourseWork"
format: html
editor: visual
---

## For the love of god don't forget

Do not display too much raw R output (e.g. don’t display the full output of
‘summary(model)’), but edit this down to the essentials. Ensure to include justification for each step of your
analyses, providing comments alongside your R code to explain what you are doing and add appropriate titles
and labelled axes to your plots.


## Question 1

```{r tidy=TRUE, tidy.opts=list(width.cutoff=72), echo=FALSE}

load("C:/AppliedDataScienceAndStatistics/Applied-Data-Science-and-Statistics/Term2/StatisticalDataModeling/datasets_exercises_Assessment.RData")

## https://upyesp.org/posts/makrdown-vscode-math-notation/ // cheat worksheet
```

We have the model: 

$Yi$ ~ $N(\frac{\theta1xi}{\theta2 + xi}$, $\sigma^2)$

## Question 1 a)

Due to the visible non-linearity of the model, we would be required to significantly transform our data to get a linear model that would have an acceptable fit of the data. We can also see that the response data seems to be only positive while a normal distribution goes from ]-$\infty$,$\infty$[ . Such arbitrary transformation increases the complexity of the model, making it less interpretable and not respect the nature of the data.

//should I talk about the increase in variance or is it only topic 2?



## Question 1 b)



```{r tidy=TRUE, tidy.opts=list(width.cutoff=72)}



```


```{r tidy=TRUE, tidy.opts=list(width.cutoff=72)}



```

```{r}
1 + 1
```

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is
displayed).
