---
title: "AdvancedStatsAssignment"
format: pdf
editor: visual
---


```{r tidy=TRUE, tidy.opts=list(width.cutoff=72), echo=FALSE, output=FALSE}

library(C50)
library(caret) # function for creating train/test datasets, and confusion matrices
library(class) # this has the knn function
library(ggplot2)
library(tidyverse)
library(tidyr)
library(DataExplorer)


classificationDF = read_csv("C:/AppliedDataScienceAndStatistics/Applied-Data-Science-and-Statistics/Term2/AdvancedTopicsInStatistics/AssignmentData/Classification.csv")

```

## Question 1



```{r tidy=TRUE, tidy.opts=list(width.cutoff=72)}



```

## Question 2 - classification

## Question 2.1

Group 0 is characterised for always having x1 > -1,65 and x2 < 2,05 with the majority of the occurences of group 0 having 

```{r tidy=TRUE, tidy.opts=list(width.cutoff=72)}

## I am just going to split the dataframe in two so its easier to show an histogram of each variable for both groups

group0 <- subset(classificationDF, Group == 0)
group1 <- subset(classificationDF, Group == 1)

##removing the id variable from both of these new data frames so they are not included in the histograms
group0 = group0 %>% select(-(1))
group1 = group1 %>% select(-(1))


plot_histogram(group0)


```
As we can see from the histograms from group 0 it seems that X2 seems to follow a normal distribution that is very slightly skewed to the left and has mean -0,5.
X1 is clearly skewed to the right possibly following a normal distribution and again mean equal to -0,5.


```{r tidy=TRUE, tidy.opts=list(width.cutoff=72)}

plot_histogram(group1)


```
Group 1 histograms reveal that X2 is normally distributed with mean equal to 0,5 and sightly skewed to the left.
The variable X1 also seems to follow a normal distribution that has a very prolonged skewness to the left and mean 0,5.


## Question 2.2 

```{r tidy=TRUE, tidy.opts=list(width.cutoff=72)}


```

```{r tidy=TRUE, tidy.opts=list(width.cutoff=72)}



```
## Question 2.3

```{r tidy=TRUE, tidy.opts=list(width.cutoff=72)}



```



```{r tidy=TRUE, tidy.opts=list(width.cutoff=72)}



```



```{r tidy=TRUE, tidy.opts=list(width.cutoff=72)}



```

```{r tidy=TRUE, tidy.opts=list(width.cutoff=72)}



```