---
title: "Class1StatsTopics"
format:
  pdf:
    toc: true
editor: visual
---

## Quarto

```{r tidy=TRUE, tidy.opts=list(width.cutoff=72)}
# install.packages('R2jags', dependencies = TRUE) # installs coda too
# # install.packages('coda', dependencies = TRUE)
# install.packages('lattice',dependencies = TRUE)
# install.packages('MCMCvis', dependencies = TRUE)
```

```{r tidy=TRUE, tidy.opts=list(width.cutoff=72)}
library(R2jags); library(MCMCvis); library(coda); library(lattice)
```

```{r tidy=TRUE, tidy.opts=list(width.cutoff=72)}
#creating a sample o n size to fit out posterior distribution
#sample(x, size,...)

```

## Coin example

```{r tidy=TRUE, tidy.opts=list(width.cutoff=72)}

# model definition
jags.mod.coin <- function(){
Y ~ dbin(0.5,10) # our data model
P8 <- ifelse(Y>7,1,0) # the probability of interest, aka 8 or higher out of 10
## the ifelse gives the first value if the statement is true and the second value if the statement is false.
## we are not forced to use the binary 1 and 0, we can use any value
}
## jag has 2 syntaxes
## the first one is a stochastic dependence ~
## the second one is a logical dependence <-

##in the jag binom we use the mu parameter and the tau parameter
## the tau parameter is called the precision and is 1/ sigma(σ) ^2

```

```{r tidy=TRUE, tidy.opts=list(width.cutoff=72)}

jags.mod.fit.coin <- jags(data = list(), model.file = jags.mod.coin,
parameters.to.save = c('Y','P8'),n.chains=1,
DIC=FALSE, n.burnin=0,n.iter = 100)

```
To get the numerical summary of the above model run we use the print function
```{r tidy=TRUE, tidy.opts=list(width.cutoff=72)}

print(jags.mod.fit.coin)

```
We can also plot the simulated samples using the traceplot function
```{r tidy=TRUE, tidy.opts=list(width.cutoff=72)}

traceplot(jags.mod.fit.coin)

```
For more diagnostics and visualisation tools we can convert the output of the jags function into an MCMC
object. Using the MCMC object we can look at numerical summaries, traceplots and density plots.
```{r tidy=TRUE, tidy.opts=list(width.cutoff=72)}

# convert into MCMC object
jagsfit.mcmc.coin <- as.mcmc(jags.mod.fit.coin)
# get numerical summary
summary(jagsfit.mcmc.coin)

```

```{r tidy=TRUE, tidy.opts=list(width.cutoff=72)}

# get traceplots
xyplot(jagsfit.mcmc.coin)
# get density estimate
densityplot(jagsfit.mcmc.coin)

```

```{r tidy=TRUE, tidy.opts=list(width.cutoff=72)}

MCMCtrace(jagsfit.mcmc.coin,
params = 'Y', # parameter of interest
type = 'density', # density plot
ind = TRUE, # separate density lines for each chain
pdf = FALSE) # plots are NOT exported into a pdf


MCMCtrace(jagsfit.mcmc.coin,
params = 'P8',
type = 'trace',
ind = TRUE,
pdf = FALSE)

```

```{r tidy=TRUE, tidy.opts=list(width.cutoff=72)}

jags.mod.fit.coin$BUGSoutput$summary[,1] # mean
jags.mod.fit.coin$BUGSoutput$summary[,3] # 2.5 percentile
jags.mod.fit.coin$BUGSoutput$summary[,7] # 97.5 percentile

```
## Exercise 1
```{r tidy=TRUE, tidy.opts=list(width.cutoff=72)}

##increasing the number of iterations
## the mean estimate is much closer to the truth, more reliable estimates
jags.mod.fit.coin <- jags(data = list(), model.file = jags.mod.coin,
parameters.to.save = c('Y','P8'),n.chains=1,
DIC=FALSE, n.burnin=0,n.iter = 100000)


print(jags.mod.fit.coin)


```
## Exercise 2
```{r tidy=TRUE, tidy.opts=list(width.cutoff=72)}

##number of chains, how much times does jags runs the simulation independently of each other

jags.mod.fit.coin <- jags(data = list(), model.file = jags.mod.coin,
parameters.to.save = c('Y','P8'),n.chains=3,
DIC=FALSE, n.burnin=0,n.iter = 50)


# convert into MCMC object
jagsfit.mcmc.coin <- as.mcmc(jags.mod.fit.coin)
# get numerical summary
summary(jagsfit.mcmc.coin)

# get trace plots
xyplot(jagsfit.mcmc.coin)
# get density estimate
densityplot(jagsfit.mcmc.coin)

```
## Exercise 3

```{r tidy=TRUE, tidy.opts=list(width.cutoff=72)}

#model definition
jags.mod.coin <- function(){
Y ~ dbin(0.5,10) # our data model
P8 <- ifelse(Y>7,1,0) # the probability of interest, aka 8 or higher out of 10
}

```

## Exercise 4
```{r tidy=TRUE, tidy.opts=list(width.cutoff=72)}

#model definition with var 1 and sd 2
jags.norm3 <- function(){
Y ~ dnorm(1,1/4) # 1/var=1 1/sd^2= 1/4
x = Y^3
}


jags.mod.fit.coin <- jags(data = list(), model.file = jags.norm3,
parameters.to.save = c('Y','P8'),n.chains=2,
DIC=FALSE, n.burnin=0,n.iter = 5000)

# convert into MCMC object
jagsfit.mcmc.coin <- as.mcmc(jags.mod.fit.coin)
# get numerical summary
#summary(jagsfit.mcmc.coin)

##need to change the entry variable
# get trace plots
xyplot(jagsfit.mcmc.coin)
# get density estimate
densityplot(jagsfit.mcmc.coin)



```


```{r tidy=TRUE, tidy.opts=list(width.cutoff=72)}



```


```{r tidy=TRUE, tidy.opts=list(width.cutoff=72)}



```


```{r tidy=TRUE, tidy.opts=list(width.cutoff=72)}



```


```{r tidy=TRUE, tidy.opts=list(width.cutoff=72)}



```


```{r tidy=TRUE, tidy.opts=list(width.cutoff=72)}



```
