---
title: "Comunicating data science"
format:
  pdf:
    toc: true
    toc-depth: 4
    # number-sections: true
editor: visual
bibliography: cod.bibtex
---

## I. Introduction {#sec-introduction1}

<!-- I. Introduction -->
<!-- A. Background and context -->
<!-- B. Purpose and objectives of the report -->
<!-- C. Scope and limitations of the report -->

### Introduction {#sec-introduction2}


Atlantic cod also known by its scientific name Gadus morhua is one of the most commercially important fish species in the world, with a long history of human exploitation dating back to the 10th century during by the Vikings @Kurlansky1998Cod. 

This species has been an important source of food and income for coastal communities across the North Atlantic region, supporting a vast industry of fishermen, processors, and distributors for centuries. However, the sustainability of the Atlantic cod fishery has been increasingly threatened in recent years due to overfishing, habitat destruction, and climate change.

The warming of ocean waters, caused by increasing greenhouse gas emissions, has altered the biology, ecology, and behaviour of Atlantic cod, leading to declines in population and distribution. These changes have had significant impacts on the commercial fishing industry, which has relied heavily on Atlantic cod as a source of income and food for centuries. 

In recent years, scientific research and policy efforts have been focused on understanding the impacts of climate change on Atlantic cod, and developing sustainable fishing strategies that take into account the changing environmental conditions. 

This report aims to review the current state of knowledge on the impact of climate change on Atlantic cod, including the ecological and environmental factors that shape its distribution and abundance, considering some of the socio-economic implications of its decline, and the strategies for sustainable management and conservation in the face of climate change.

Through an analysis of the data regarding sea surface temperature increase and fishing, this report seeks to contribute to a more comprehensive and evidence-based understanding of the challenges associated with the Atlantic cod fishery in the era of climate change.

## Background and context

Atlantic cod fishing has a long and complex history that dates back to at least the 11th century. Cod was a vital food source for European colonizers and became one of the most important commercial fish species in the North Atlantic.

The first recorded European explorers to fish for cod in North America were the Vikings, who fished for cod off the coast of Newfoundland around 1000 AD. Later, in the 15th and 16th centuries, European fishermen established seasonal fishing camps along the New England nicknaming it Cape Cod @Kurlansky1998CodP63.

As European fishing fleets grew in size and range, they began to explore deeper waters and more distant fishing grounds, leading to the development of large-scale commercial fishing operations. By the late 18th century, French and English fleets were competing for the rich cod fisheries of the Grand Banks and other areas leading to the modernization and a significant increase of cod fishing @Kurlansky1998CodP118.

As such from its beginning of being an important food source, to its increased values in trade as a commodity and its increased in significance culturally due to it becoming a staple of many coastal diets and the naval development that produced experienced seamen and naval infrastructure that was required during the naval arms race of the colonial era.


## Objectives

The report aims to analyse the impacts of climate change on Atlantic cod fisheries in  Iceland, Norway, Denmark and the United Kingdom. The objectives of the report include evaluating the state of Atlantic cod fisheries from 1950 to 2010, identifying the scientific evidence linking climate change to changes in Atlantic cod's biology, ecology, and behaviour, identifying the key challenges facing the fisheries, and examining potential innovative management approaches to promote the resilience, adaptation, and sustainability of Atlantic cod fisheries.

Overall, the report aims to provide a comprehensive analysis of the ecological consequences of climate change on Atlantic cod fisheries and propose strategies for adaptation and mitigation.

## Scope and limitations

The report is limited in scope to the impacts of climate change on Atlantic cod fisheries in Iceland, Norway, Denmark and Great Britain from 1950 to 2010. The report does not cover other fish species or historical changes in Atlantic cod populations prior to the onset of climate change. The analysis is based on existing scientific research and data, which may be limited in scope or availability. While the report provides an analysis of the ecological impact of climate change on Atlantic cod fisheries, it does not advocate for specific management strategies or policy recommendations. The report aims to provide a comprehensive analysis of the issues and potential solutions within the knowledge and expertise limitations of the author.

## TODO talk here about overfishing as a limitation that is hard to account for


## II. Climate Change and its Impact on Cod Fisheries

<!-- A. Causes and Effects of Climate Change -->
<!-- B. Historical Changes in Cod Populations -->
<!-- C. Current Situation of Cod Fisheries -->

## Causes and Effects of Climate Change
Climate change is primarily caused by the increase in greenhouse gases (GHGs) in the atmosphere, particularly carbon dioxide (CO2). Human activities, such as burning fossil fuels, deforestation, and agriculture, are the main sources of GHG emissions. These activities release large amounts of CO2 and other GHGs into the atmosphere, trapping heat and warming the planet.

## Causes and Effects of Climate Change

III. The Effects of Climate Change on Cod Fisheries in Denmark, Iceland, and Norway
A. Denmark
Overview of Denmark's Cod Fisheries
The Effects of Climate Change on Denmark's Cod Fisheries
B. Iceland
Overview of Iceland's Cod Fisheries
The Effects of Climate Change on Iceland's Cod Fisheries
C. Norway
Overview of Norway's Cod Fisheries
The Effects of Climate Change on Norway's Cod Fisheries
C. Scotland
Overview of Scotland's Cod Fisheries
The Effects of Climate Change on Scotland's Cod Fisheries
    
V. Conclusion
A. Summary of Findings
B. Implications for Cod Fisheries and Fishing Communities
C. Future Directions for Research and Action

<!-- VI. References -->
<!-- A. List of Sources Cited -->
<!-- B. Additional Resources for Further Reading -->


```{r tidy=TRUE, tidy.opts=list(width.cutoff=72), echo=FALSE, warning=FALSE, output=FALSE}

library(ggplot2)
library(tidyverse)
library(tidyr)
library(DataExplorer)

library(readr)
library(lubridate)

library(tseries)
library(forecast)

#install.packages("forecast")


FishDataSet = read_csv("HistoricalLandings1950-2010/ICES_1950-2010.csv")

## https://www.st.nmfs.noaa.gov/copepod/toolkit/subform_copepodite-c2.html
## get the data from a location

```



```{r tidy=TRUE, tidy.opts=list(width.cutoff=72), echo=FALSE, warning=FALSE, output=FALSE}
#FishDataSet$Species

BacalhauDataFrame <- subset(FishDataSet, Species == "Atlantic cod")

```

```{r tidy=TRUE, tidy.opts=list(width.cutoff=72), echo=FALSE, warning=FALSE, output=FALSE}
#BacalhauDataFrame$Country

BomBacalhauDataFrame <- subset(BacalhauDataFrame, Country %in% c("Norway", "Iceland", # "Greenland",
  "Denmark","Faeroe Islands", "UK - England & Wales", "UK - Scotland" ))
```



```{r tidy=TRUE, tidy.opts=list(width.cutoff=72), echo=FALSE, warning=FALSE, output=FALSE}

# total_count <- sum(BomBacalhauDataFrame$1951)

BomBacalhauDataFrameNumeric <- BomBacalhauDataFrame %>% 
     mutate_at(c(4:64), as.numeric)

BomBacalhauDataFrameNumeric = BomBacalhauDataFrameNumeric %>% select(-(Species:Division))

BomBacalhauDataFrameNumeric = BomBacalhauDataFrameNumeric %>% replace(is.na(.), 0)


# Summarize all selected columns by summing them
BacalhauSomado<- BomBacalhauDataFrameNumeric %>%
  group_by(Country) %>%
  summarise_at(vars(`1950`:`2010`), sum, na.rm = TRUE)


## now I just need to see the sea surface temperature change from 1950 to 2010 


```

```{r tidy=TRUE, tidy.opts=list(width.cutoff=72), echo=FALSE, warning=FALSE, output=FALSE}

IcelandSeaSurfaceTemperature = read_csv("SeaSurfaceTemperature/IcelandSeaSurfaceTemperature.csv")

NorwaySeaSurfaceTemperature = read_csv("SeaSurfaceTemperature/NorwaySeaSurfaceTemperature.csv")

DenmarkSeaSurfaceTemperature = read_csv("SeaSurfaceTemperature/DenmarkSeaSurfaceTemperature.csv")

FaroeIslandsSeaSurfaceTemperature = read_csv("SeaSurfaceTemperature/FaroeIslandsSeaSurfaceTemperature.csv")

#EnglandWalesSeaSurfaceTemperature = read_csv("SeaSurfaceTemperature/EnglandWalesSeaSurfaceTemperature.csv")

ScotlandSeaSurfaceTemperature = read_csv("SeaSurfaceTemperature/ScotlandSeaSurfaceTemperature.csv")


```

```{r tidy=TRUE, tidy.opts=list(width.cutoff=72), echo=FALSE, warning=FALSE, output=FALSE}

## removing the third non existent variable form all the datasets

IcelandSeaSurfaceTemperature = IcelandSeaSurfaceTemperature %>% select(-(...3))
NorwaySeaSurfaceTemperature = NorwaySeaSurfaceTemperature %>% select(-(...3))
DenmarkSeaSurfaceTemperature = DenmarkSeaSurfaceTemperature %>% select(-(...3))
FaroeIslandsSeaSurfaceTemperature = FaroeIslandsSeaSurfaceTemperature %>% select(-(...3))
ScotlandSeaSurfaceTemperature = ScotlandSeaSurfaceTemperature %>% select(-(...3))

##maybe add england + wales if needed

str(IcelandSeaSurfaceTemperature)
## the data was indeed correctly stored in the date YYYY-MM-DD format

```

```{r tidy=TRUE, tidy.opts=list(width.cutoff=72), echo=FALSE, warning=FALSE, output=FALSE}

IcelandSeaSurfaceTemperature <- IcelandSeaSurfaceTemperature[-c(1:600), ]

IcelandSeaSurfaceTemperature <- IcelandSeaSurfaceTemperature[-c(733:864), ]

NorwaySeaSurfaceTemperature <- NorwaySeaSurfaceTemperature[-c(1:600), ]

NorwaySeaSurfaceTemperature <- NorwaySeaSurfaceTemperature[-c(733:864), ]

FaroeIslandsSeaSurfaceTemperature <- FaroeIslandsSeaSurfaceTemperature[-c(1:600), ]

FaroeIslandsSeaSurfaceTemperature <- FaroeIslandsSeaSurfaceTemperature[-c(733:864), ]

ScotlandSeaSurfaceTemperature <- ScotlandSeaSurfaceTemperature[-c(1:600), ]

ScotlandSeaSurfaceTemperature <- ScotlandSeaSurfaceTemperature[-c(733:864), ]

```

```{r tidy=TRUE, tidy.opts=list(width.cutoff=72), echo=FALSE, warning=FALSE, output=FALSE}

# Create time series object
timeSeriesIceland <- ts(IcelandSeaSurfaceTemperature$`HSST= HadISST Temperature (C)` , start = ymd("1950-01-15"), frequency = 12)

plot(timeSeriesIceland, main = "Temperature Time Series")

```

```{r tidy=TRUE, tidy.opts=list(width.cutoff=72)}

adf.test(timeSeriesIceland)

```
If the p-value of the test is less than 0.05, you can reject the null hypothesis that the time series is non-stationary and assume that it is stationary.

```{r tidy=TRUE, tidy.opts=list(width.cutoff=72)}

arima_model <- auto.arima(timeSeriesIceland)

summary(arima_model)

```

```{r tidy=TRUE, tidy.opts=list(width.cutoff=72)}


```

```{r tidy=TRUE, tidy.opts=list(width.cutoff=72)}


```



## References
