---
title: "Comunicating data science"
format:
  pdf:
    toc: true
editor: visual
---

## Introduction

I. Introduction
A. Background and context
B. Purpose and objectives of the report
C. Scope and limitations of the report

II. Climate Change and its Impact on Cod Fisheries
A. Causes and Effects of Climate Change
B. Historical Changes in Cod Populations
C. Current Situation of Cod Fisheries

III. The Effects of Climate Change on Cod Fisheries in Greenland, Iceland, and Norway
A. Greenland

    Overview of Greenland's Cod Fisheries
    The Effects of Climate Change on Greenland's Cod Fisheries
    B. Iceland
    Overview of Iceland's Cod Fisheries
    The Effects of Climate Change on Iceland's Cod Fisheries
    C. Norway
    Overview of Norway's Cod Fisheries
    The Effects of Climate Change on Norway's Cod Fisheries
    
V. Conclusion
A. Summary of Findings
B. Implications for Cod Fisheries and Fishing Communities
C. Future Directions for Research and Action

VI. References
A. List of Sources Cited
B. Additional Resources for Further Reading


```{r tidy=TRUE, tidy.opts=list(width.cutoff=72), echo=FALSE, warning=FALSE, output=FALSE}

library(ggplot2)
library(tidyverse)
library(tidyr)
library(DataExplorer)


FishDataSet = read_csv("C:/AppliedDataScienceAndStatistics/Applied-Data-Science-and-Statistics/Term2/CommunicatingDataScience/IndividualReport/HistoricalLandings1950-2010/ICES_1950-2010.csv")

```



```{r tidy=TRUE, tidy.opts=list(width.cutoff=72), echo=FALSE}
#FishDataSet$Species

BacalhauDataFrame <- subset(FishDataSet, Species == "Atlantic cod")

```

```{r tidy=TRUE, tidy.opts=list(width.cutoff=72), eval=FALSE}
#BacalhauDataFrame$Country

BomBacalhauDataFrame <- subset(BacalhauDataFrame, Country %in% c("Norway", "Iceland", "Greenland"))
```

```{r tidy=TRUE, tidy.opts=list(width.cutoff=72), eval=FALSE}

BomBacalhauDataFrame <- subset(BacalhauDataFrame, Country %in% c("Norway", "Iceland", "Greenland"))


```


```{r tidy=TRUE, tidy.opts=list(width.cutoff=72), eval=FALSE}

# total_count <- sum(BomBacalhauDataFrame$1951)

BomBacalhauDataFrameNumeric <- BomBacalhauDataFrame %>% 
     mutate_at(c(4:64), as.numeric)

BomBacalhauDataFrameNumeric = BomBacalhauDataFrameNumeric %>% select(-(Species:Division))

BomBacalhauDataFrameNumeric = BomBacalhauDataFrameNumeric %>% replace(is.na(.), 0)


# Summarize all selected columns by summing them
BacalhauSomado<- BomBacalhauDataFrameNumeric %>%
  group_by(Country) %>%
  summarise_at(vars(`1950`:`2010`), sum, na.rm = TRUE)


## now I just need to see the sea surface temperature change from 1950 to 2010 


```

```{r tidy=TRUE, tidy.opts=list(width.cutoff=72)}


```

```{r tidy=TRUE, tidy.opts=list(width.cutoff=72)}


```

```{r tidy=TRUE, tidy.opts=list(width.cutoff=72)}


```

```{r tidy=TRUE, tidy.opts=list(width.cutoff=72)}


```

```{r tidy=TRUE, tidy.opts=list(width.cutoff=72)}


```

```{r tidy=TRUE, tidy.opts=list(width.cutoff=72)}


```

```{r tidy=TRUE, tidy.opts=list(width.cutoff=72)}


```

```{r tidy=TRUE, tidy.opts=list(width.cutoff=72)}


```



You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is
displayed).
